Minimal trap spaces of Logical models are maximal siphons of their Petri net encoding

### 2022-05-10 cmsb22.tex
Some small corrections

Boolean modelling has proven over the years that it can bring powerful analyses and corresponding insight to the many cases where precise biological data is not sufficiently available to build a detailed quantitative model~\cite{wang2012boolean}, even for modelling post-transcriptional mechanisms.
This is even more true for very large models where such data is frequently missing and led to a constant increase in size of logical models \emph{à la} Thomas~\cite{aghamiri2020automated}.
Besides simulation, the analysis of such models is mostly based on attractor computation, since those correspond roughly to observable biological \emph{phenotypes}. The recent use of trap spaces~\cite{klarner2015computing} made a real breakthrough in that field allowing to consider medium-sized models that used to be out of reach.
However, with the continuing increase in model-size, the state of the art computation of minimal trap spaces based on \emph{prime-implicants} shows its limits as there can be a huge number of implicants.

Petri nets were introduced in the 60’s as a simple formalism for describing and analyzing information processing systems that are characterized as being concurrent, asynchronous, non-deterministic and possibly distributed~\cite{peterson1981petri,Murata1989}.
The use of Petri nets for representing biochemical reaction systems, by mapping molecular species to places and reactions to transitions, was introduced quite late in~\cite{reddy1993petri}, together with some Petri net concepts and tools for the analysis of metabolic networks.
Siphons are such a concept but they have not been used a lot for the analysis of biochemical systems even if the practical cost of computing their minimal/maximal elements appear much more manageable than the theoretical complexity would indicate~\cite{nabli2016enumerating}.

In this article we present an alternative method to compute minimal trap spaces, and hence complex attractors, of a Boolean model. It replaces the need for prime-implicants by a completely different technique, namely the enumeration of maximal siphons in the Petri net encoding of the original model.
After some technical preliminaries, we expose the concrete need for such a method and detail its implementation using Answer Set Programming (ASP).
We then demonstrate its efficiency and compare it to implicant-based methods on some large Boolean models from the literature.

@@

\begin{table}[!htb]
  \caption{Timing comparisons between the proposed method and PyBoolNet on the PyBoolNet's repository.
    Column "\(n\)" denotes the numbers of nodes of each model. Column "\(|M|\)" denotes the number of minimal trap spaces and for each method is given the computation time in seconds.
    "DNF" means that the method did not finish the computation (with a maximum of 1000 trap spaces) within the timeout of 2 minutes. Number in bold indicate a ratio greater than three.}
  \centering
  \label{tab:pyboolnet_repo}
  \begin{tabular}{lrrrrrr}
    \toprule
    model & $n$ & $|M|$ & PyBoolNet (s) & Trappist (s) \\ \midrule
    arellano\_rootstem & 9 & 4 & 0.04 & \textbf{0.31}\\
    calzone\_cellfate & 28 & 27 & 0.03 & 0.08\\
    dahlhaus\_neuroplastoma & 23 & 32 & 0.04 & 0.11\\
    davidich\_yeast & 10 & 12 & 0.03 & 0.08\\
    dinwoodie\_life & 15 & 7 & 0.03 & 0.08\\
    dinwoodie\_stomatal & 13 & 1 & 0.04 & 0.05\\
    faure\_cellcycle & 10 & 2 & 0.04 & 0.04\\
    grieco\_mapk & 53 & 18 & 0.04 & 0.08\\
    irons\_yeast & 18 & 1 & 0.05 & 0.04\\
    jaoude\_thdiff & 103 & 1000+ & \textbf{1.73} & 0.40\\
    klamt\_tcr & 40 & 8 & 0.06 & 0.04\\
    krumsiek\_myeloid & 11 & 6 & 0.03 & 0.03\\
    multivalued & 13 & 4 & 0.02 & 0.02\\
    n12c5 & 11 & 5 & \textbf{34.37} & 0.06\\
    n3s1c1a & 2 & 2 & 0.02 & 0.03\\
    n3s1c1b & 2 & 2 & 0.02 & 0.03\\
    n5s3 & 4 & 3 & 0.03 & 0.03\\
    n6s1c2 & 5 & 3 & 0.03 & 0.04\\
    n7s3 & 6 & 3 & 0.02 & 0.02\\
    raf & 3 & 2 & 0.03 & 0.02\\
    randomnet\_n15k3 & 15 & 3 & 0.04 & 0.05\\
    randomnet\_n7k3 & 7 & 10 & 0.03 & \textbf{0.19}\\
    remy\_tumorigenesis & 34 & 25 & \textbf{1.79} & 0.13\\
    saadatpour\_guardcell & 13 & 1 & 0.02 & 0.03\\
    selvaggio\_emt & 56 & 1000+ & \textbf{1.03} & 0.23\\
    tournier\_apoptosis & 12 & 3 & 0.05 & 0.06\\
    xiao\_wnt5a & 7 & 4 & 0.06 & 0.03\\
    zhang\_tlgl & 60 & 156 & 0.26 & 0.17\\
    zhang\_tlgl\_v2 & 60 & 258 & 0.19 & 0.17\\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

Boolean modelling has proven over the years that it can bring powerful analyses and corresponding insight to the many cases where precise biological data is not sufficiently available to build a detailed quantitative model~\cite{wang2012boolean}, even for modelling post-transcriptional mechanisms.
This is even more true for very large models where such data is frequently missing and led to a constant increase in size of logical models \emph{à la} Thomas~\cite{aghamiri2020automated}.
Besides simulation, the analysis of such models is mostly based on attractor computation, since those correspond roughly to observable biological \emph{phenotypes}. The recent use of trap spaces~\cite{klarner2015computing} made a real breakthrough in that field allowing to consider medium-sized models that used to be out of reach.
However, with the continuing increase in model-size, the state-of-the-art computation of minimal trap spaces based on \emph{prime-implicants} shows its limits as there can be a huge number of implicants.

Petri nets were introduced in the 60’s as simple formalism for describing and analyzing information processing systems that are characterized as being concurrent, asynchronous, non-deterministic and possibly distributed~\cite{peterson1981petri,Murata1989}.
The use of Petri nets for representing biochemical reaction systems, by mapping molecular species to places and reactions to transitions, was introduced quite late in~\cite{reddy1993petri}, together with some Petri net concepts and tools for the analysis of metabolic networks.
Siphons are such a concept but they have not been used a lot for the analysis of biochemical systems even if the practical cost of computing their minimal/maximal elements appear much more manageable than the theoretical complexity would indicate~\cite{nabli2016enumerating}.

In this article we present an alternative method to compute minimal trap spaces, and hence complex attractors, of a Boolean model. It replaces the need for prime-implicants by a completely different technique, namely the enumeration of maximal siphons in the Petri net encoding of the original model.
After some technical preliminaries, we expose the concrete need for such a method and detail its implementation using Answer Set Programming.
We then demonstrate its efficiency and compare it to implicant-based methods on several large Boolean models from the literature.

@@

\begin{table}[!htb]
  \caption{Timing comparisons between the proposed method and PyBoolNet on the PyBoolNet's repository.
    Column "\(n\)" denotes the numbers of nodes of each model. Column "\(|M|\)" denotes the number of minimal trap spaces and for each method is given the computation time in seconds.
    "DNF" means that the method did not finish the computation (with a maximum of 1000 trap spaces) within the timeout of 2 minutes. A number in bold indicates a ratio greater than three.}
  \centering
  \label{tab:pyboolnet_repo}
  \begin{tabular}{lrrrrrr}
    \toprule
    model & $n$ & $|M|$ & PyBoolNet (s) & Trappist (s) \\ \midrule
    arellano\_rootstem & 9 & 4 & 0.04 & \textbf{0.31}\\
    calzone\_cellfate & 28 & 27 & 0.03 & 0.08\\
    dahlhaus\_neuroplastoma & 23 & 32 & 0.04 & 0.11\\
    davidich\_yeast & 10 & 12 & 0.03 & 0.08\\
    dinwoodie\_life & 15 & 7 & 0.03 & 0.08\\
    dinwoodie\_stomatal & 13 & 1 & 0.04 & 0.05\\
    faure\_cellcycle & 10 & 2 & 0.04 & 0.04\\
    grieco\_mapk & 53 & 18 & 0.04 & 0.08\\
    irons\_yeast & 18 & 1 & 0.05 & 0.04\\
    jaoude\_thdiff & 103 & 1000+ & \textbf{1.73} & 0.40\\
    klamt\_tcr & 40 & 8 & 0.06 & 0.04\\
    krumsiek\_myeloid & 11 & 6 & 0.03 & 0.03\\
    multivalued & 13 & 4 & 0.02 & 0.02\\
    n12c5 & 11 & 5 & \textbf{34.37} & 0.06\\
    n3s1c1a & 2 & 2 & 0.02 & 0.03\\
    n3s1c1b & 2 & 2 & 0.02 & 0.03\\
    n5s3 & 4 & 3 & 0.03 & 0.03\\
    n6s1c2 & 5 & 3 & 0.03 & 0.04\\
    n7s3 & 6 & 3 & 0.02 & 0.02\\
    raf & 3 & 2 & 0.03 & 0.02\\
    randomnet\_n15k3 & 15 & 3 & 0.04 & 0.05\\
    randomnet\_n7k3 & 7 & 10 & 0.03 & \textbf{0.19}\\
    remy\_tumorigenesis & 34 & 25 & \textbf{1.79} & 0.13\\
    saadatpour\_guardcell & 13 & 1 & 0.02 & 0.03\\
    selvaggio\_emt & 56 & 1000+ & \textbf{1.03} & 0.23\\
    tournier\_apoptosis & 12 & 3 & 0.05 & 0.06\\
    xiao\_wnt5a & 7 & 4 & 0.06 & 0.03\\
    zhang\_tlgl & 60 & 156 & 0.26 & 0.17\\
    zhang\_tlgl\_v2 & 60 & 258 & 0.19 & 0.17\\
    \bottomrule
  \end{tabular}
\end{table}
